---
title: 'SmarteR not HardeR'
subtitle: 'R in the Psychological Science Production Cycle'
author: "Matthew Barstead, Ph.D."
date: "11/24/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, tidy.opts = list(width.cutoff=60))
```

## Overview
- Brief Overview of `R`
- RStudio and Getting the Most from `R`
- `R` objects
- Sample Size Planning
- Exploratory Data Analysis
- MLM Basics
- Data Processing, Documentation, and Management

## Who am I? 
- Received my Ph.D. from the Department of Human Development and Quantitative Methodology at UMD
- Primary advisor Ken Rubin - child and adolescent social and emotional development
    * Heavy focus on observational methodologies (behavioral coding schemes)
- Dissertation focused on the creation of an RShiny application for processing and editing heart rate data
- Currently a data scientist/software engineer at Capital One
    * Work on the analytics and infrastructure team for U.S. Consumer Card 
    
## Me and R
- First real exposure in 2011 and mainly for power analyses to start
- Migrated nearly all modeling over to `R` during coursework 
    - GLMs $\checkmark$
    - Multilevel Models $\checkmark$
    - Structural Equation Models $\checkmark$
    - Bayesian extensions via BUGS $\checkmark$, JAGS $\checkmark$, Stan $\checkmark$
    - Unsupervised learning $\checkmark$
    - Timeseries, signal processing, imputation models... you get the picture. 

## Brief Overview of R 
**History**  
`R` is an implementation of the `S` programming language. The intent was to make a more interactive language for rapid development and deployment of statistical models. Two researchers from the University of Auckland (**R**oss Ihaka and **R**obert Gentleman), are responsible for creating the first version of `R`.  

## `R` vs. `Python` 
Both are sometimes referred to as object-oriented programming languages. Only *really* true of `Python`. 

`Python`:
```{python, cars+_py, eval=FALSE}
cars.describe()
```
`R`: 
```{r, cars_r}
summary(cars)
```

## Getting the Most From `R`
- [RStudio](https://rstudio.com/products/rstudio/download/) 
- [RMarkdown](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [RShiny](https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)
- [`ggplot`](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`plotly`](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf)

## `R` Objects
**`character` (i.e., strings)** 

```{r, char_Ex1}
char_Ex <- "This is a string object in R"
class(char_Ex)
```

## `R` Objects
**`character` (i.e., strings)** 

What if I wanted to get all of the words in `char_Ex` individually in a separate object?
```{r, char_Ex2}
strsplit(char_Ex, split=" ")
```

For more complex `character` processing examples see [this site](https://www.mjdenny.com/Text_Processing_In_R.html), which relies heavily on the `stringr` package in `R`

## `R` Objects
**`character` (i.e., strings)**

The `paste()` and `paste0()` functions are among the most useful. 

```{r}
max <- 3
for(i in 1:max){
    # some functional code goes here:
    msg <- paste("Simulation", i, "of", max, "completed")
    print(msg)
}
```

## `R` Objects

Numeric objects: `intergers` and `numeric` "types" 

```{r}
int_val <- integer(1)
paste("Object class type:", class(int_val))
paste("Test int_val is numeric:", is.numeric(int_val))
paste("Test int_val is integer:", is.integer(int_val))
```

## `R` Objects

Numeric objects: `intergers` and `numeric` "types" 

```{r}
num_val <- numeric(1)
paste("Object class type:", class(num_val))
paste("Test num_val is numeric:", is.numeric(num_val))
paste("Test num_val is integer:", is.integer(num_val))
```

## `R` Objects

Numeric objects: `intergers` and `numeric` "types" 

```{r}
def_val <- 1
paste("Object class type:", class(def_val))
paste("Test def_val is numeric:", is.numeric(def_val))
paste("Test def_val is integer:", is.integer(def_val))
```


## `R` Objects
**`functions`**

A quick diversion for code hygiene. Notice I did a lot of copying and pasting there. As a rule of thumb, it is best to write custom functions for these sorts of situations. Less code to maintain overall. 

```{r, func1}
eval_data_type <- function(x){
    # The multiple print() statements ensure each prints in the console as program runs. 
    x_name <- deparse(substitute(x))
    print(paste("Object class type:", class(x)))
    print(paste("Test", x_name, "is numeric:", is.numeric(x)))
    print(paste("Test", x_name, "is integer:", is.integer(x)))
}
```

## `R` Objects
**`functions`**

```{r}
eval_data_type(int_val)
```

```{r}
eval_data_type(def_val)
```

## `R` Objects 
**`vectors`**
```{r}
vec_Ex <- c(1, 2, 3, 2, NA, 3)
length(vec_Ex)
mean(vec_Ex, na.rm = TRUE)
mean(vec_Ex)
```

## `R` Objects 
**`matrices`**
```{r}
matrix_Ex <- matrix(c(1,2,2,
                      3,2,1,
                      NA,0,19), 
                    ncol = 3,
                    byrow = TRUE)
matrix_Ex
```

For quick summaries: `colMeans()`, `colSums()`, `rowMeans()`, & `rowSums()`. Combine with logical tests to get quick useful information. For instance what happens when you try running: `colSums(is.na(matrix_Ex))`?

## `R` Objects

`data.frames` (and `lists`)

`data.frames` are essentially specially constructed `lists`. 

```{r}
df <- data.frame(var_name1 = sample(1:5, size=20, replace = TRUE),
                 var_name2 = rnorm(20, mean=50, sd=10))
dim(df)
head(df,3)
```

## `R` Objects
**`factors`** 

Both super useful and super annoying. Store values as integers with character labels referred to as `levels`. Certain functions will not accept `character` variables. Some accept `character` or `factor` variables (but may cast `character` variables into `factor` variables).

For more advanced manipulation of `factor` variables, take a look at `forcats::fct_relevel()` and the built-in. `levels()`

## Quick Review of Concepts
- `R` is more accurately described as a pseudo-object-oriented programming language
- Combine with RStudio, RMarkdown, & RShiny to get the most out of `R`
- `character`, `numeric`, `integer`, and `factor` variables each have different properties
- **DRY** code is better than **WET** code
- `vectors`, `matrices`, `data.frames`, and `lists` are the foundation of data storage in `R`

## Sample Size Planning
- The flexibility of SEM for power analyses - YAY!!
- Flexibility means more thought up front - BOO!! 
- *Not* a course on structural equation models 
    * Modeling the covariance matrix (and sometimes the mean vector)
    * Specify a series of structural paths hypothesized to account for the covariance matrix
    * Workflow using a simple regression model with path tracing 

## Summary of Sample Size Planning
- Simulations are often the most flexible, but most time-intensive to setup
- Goal is to provide a defensible analysis at determining sample size 
- When using existing tools be aware that many make default choices that may or may not align with your use-case
- For multilevel power analyses, both the `lavaan` + `simsem` and `simr` approaches would work

## Exploratory Data Analysis
To the script files!

## Summary of EDA 
- Dimensions vs. clusters
- Interpretation will change depending on the inputs 
- Consider possibility of holdout approaches whenever possible 
    * k-fold cross-validation 
    * Holdout test samples
- Utilize explore/confirm analytic approaches (but not on same sample)

## MLMs in R 
- `lme4`, `brms`, and `lavaan` can handle this models in varying ways 
- `lme4` introduced some very very simple syntax for writing these models: 

```{r, evalFALSE}
fit_mlm <- glmer(y ~ 1 + x1 + x2*cat1 + 
                     (1 + x1 + x2 | ID), 
                 data = data_set)
```

## MLMs in `lme4`
To the script files!

## Summary of MLMs
**Lots we did not cover:** 
- Fully Bayesian approach using `Stan` and `brms` 
- SEM extensions in `lavaan` (multigroup SEM, multilevel factor analysis, growth models)
- Partitioning variance (see Rights and Sterba, 2018, *Psychological Methods*)

## The Power of RMarkdown
- Generating output 
- Write your manuscript from start to finish (`papaja`) and combine with GH for version control
- Generate comprehensive and reproducible documentation 
- Create entire websites (`blogdown`) or online publications/manuals (`bookdown`)

## GitHub, R, and the ZenHub Board 
- An imperfect example, but imagine for a collaborative writing project

## Tips
- Documentation is usually good, but fully understanding it can take some getting used to. 
- `?function`, `vignette()`
- Someone has usually already solved the problem you are trying to solve
- Checkout RStudio cheatsheets, anything written by Hadley Wickham or Yihui Xie for ways to get the most out of `R`
- [The RGraph Gallery](https://www.r-graph-gallery.com/) has lots of inspiring visualizations